---
title: "Group Project 2"
author: "Nikolay Skryabin, Torsten Dane Bjerling, ...? "
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
---
```{r setup, include = FALSE}

#### Load necessary packages ####
packages <- c("knitr", "kableExtra", "magrittr", "readr", "tidyverse",
              "dplyr", "ggplot2", "tidyr","stringr", "leaflet")

install_me <- packages[!(packages %in% installed.packages()[, "Package"])]
if (length(install_me)) install.packages(install_me)

library(knitr)
library(kableExtra)
library(magrittr)
library(readr)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(leaflet)
```

## Background
The World Health Organization has recently employed a new data science initiative, *CSIT-165*, that uses data science to characterize pandemic diseases. 
*CSIT-165* disseminates data driven analyses to global decision makers.

*CSIT-165* is a conglomerate comprised of two fabricated entities: *World Health Organization (WHO)* and *U.S. Pandemic Response Team (USPRT)*. Your and your partner’s role is to play a data scientist from one of these two entities. Discuss with your partner to decide who will be part of *WHO* and *USPRT*.

## Data
> [2019 Novel Coronavirus COVID-19 (2019-nCoV) Data Repository by John Hopkins CSSE](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series)

Data for 2019 Novel Coronavirus is operated by the John Hopkins University Center for Systems Science and Engineering (JHU CSSE).
Data includes daily time series CSV summary tables, including confirmations, recoveries, and deaths. 
Country/region are countries/regions hat conform to World Health Organization (WHO).
Lat and Long refer to coordinates references for the user. 
Date fields are stored in MM/DD/YYYY format.

#### GitHub Repository of the Project
> [R-group-project-2](https://github.com/nick404s/R-group-project-2)


```{r}
# URLs for the global confirmed and death cases files 
URL_CONFIRMED_GLOBAL <- paste0("https://raw.githubusercontent.com/CSSEGISandData/",
                "COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/",
                "time_series_covid19_confirmed_global.csv")


URL_DEATHS_GLOBAL <- paste0("https://raw.githubusercontent.com/CSSEGISandData/",
                "COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/",
                "time_series_covid19_deaths_global.csv")

# URLs for the US confirmed and death cases files
URL_CONFIRMED_US <- paste0("https://raw.githubusercontent.com/CSSEGISandData/",
                "COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/",
                "time_series_covid19_confirmed_US.csv")

URL_DEATHS_US <- paste0("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/",
                     "master/csse_covid_19_data/csse_covid_19_time_series/",
                     "time_series_covid19_deaths_US.csv")

# Loads a file from URL and reads data from the file into data frame
getDataFrame <- function(url)
{
   # split the URL into list using "/" as delimiter
  url_list<-strsplit(url,split = "/")
  
  file_name <- sapply(url_list, tail, 1) # get the file name 

  download.file(url, file_name) # download the file

  # create data frame from the file
  data_frame <- read.csv(file_name, stringsAsFactors = FALSE) 

  data_frame # return it
}

# Get the data frames for the global deaths and confirmed cases
global_confirmed_df <- getDataFrame(URL_CONFIRMED_GLOBAL)
global_deaths_df <- getDataFrame(URL_DEATHS_GLOBAL)

# Get the data frames for the US deaths and confirmed cases
# us_confirmed_df <- getDataFrame(URL_CONFIRMED_US)
# us_deaths_df <- getDataFrame(URL_DEATHS_US)

```
## Project Objectives

### Objective 1
```{r }
# Calculates a sum of cases for each country
calculateSumCases <- function(df)
{
  # get max sum state using piping
  result_sum <-na.omit(df) %>%                                # take the data frame
          filter((Lat & Long) != 0, 
                 !str_detect(Country.Region, 'Olympics'),
                 !str_detect(Country.Region, 'Antarctica')) %>%
          rename(Last_Day = last_col()) %>%         # rename the last column
          group_by(Country.Region) %>%              # aggregate the states
          summarize("Sum"=sum(Last_Day,na.rm=T),
                    Lat.Mean=mean(Lat),
                    Long.Mean=mean(Long))%>% # get a sum of cases for each country
          select(Country.Region,Lat.Mean,Long.Mean,Sum) # get the country data
}

# Get the global dataframes for the confirmed and death cases data sets 
global_confirmed_sums <- calculateSumCases(global_confirmed_df) %>%
                          rename(Confirmed=Sum)

global_deaths_sums <- calculateSumCases(global_deaths_df) %>%
                      rename(Deaths=Sum)
# Add a column with the death cases to get dataframe for all cases
global_all_sums <- global_confirmed_sums %>%
                    mutate(Deaths = global_deaths_sums$Deaths)

# Converts a column numbers to formatted strings
format_numbers<- function(a_column)
{
  a_column <- paste0("",format((as.numeric(a_column)),
                               big.mark = ",", nsmall = 1))
}

# Get the colors for the markers using colorQuantile()
qpal_confirmed <- colorQuantile(c("#f59898","#fa4848","#330101"), global_all_sums$Confirmed, n = 5)

qpal_deaths <- colorQuantile(c("#0084F7","#E0E0E0","#D00000"), global_all_sums$Deaths, n = 5)

# Create a map for the confirmed and death cases
global_map <- leaflet(data = global_all_sums) %>%
  addProviderTiles("CartoDB.Positron") %>%
  setView(0, 0,  zoom = 1) %>% 
  addCircleMarkers(lat = ~Lat.Mean, lng = ~Long.Mean, 
                   popup = ~paste0("Confirmed: ",format_numbers(Confirmed)),
                   label = ~Country.Region,
                   color = ~qpal_confirmed(Confirmed), opacity = 0.7,group = "Confirmed",stroke=FALSE) %>%
  addCircleMarkers(lat = ~Lat.Mean, lng = ~Long.Mean,
                   popup = ~paste0("Deaths: ",format_numbers(Deaths)),
                   label = ~Country.Region ,color = ~qpal_deaths(Deaths), group = "Deaths",stroke=FALSE) %>%
  addLayersControl(overlayGroups = c("Confirmed","Deaths"), options = layersControlOptions(collapsed = FALSE))

print(global_map)

```



### Objective 2
```{r }
# Sorts a data by number of cases
getSortedByCases<- function(df)
{
  cases <- df[,4]
  result <- df%>%
            arrange(desc(cases)) %>%
            mutate_at(4,format_numbers)%>%
            select(Country=1, Counts=4) 
}

# Get sorted countries by confirmed cases and add a rank column
global_confirmed_rank <-getSortedByCases(global_confirmed_sums) %>%
                        mutate(Rank = row_number(), .before= Country)

# Get sorted countries by death cases
global_deaths_rank <-getSortedByCases(global_deaths_sums)

# Display 2 tables with confirmations and deaths
kbl(list(global_confirmed_rank,global_deaths_rank)) %>%
  kable_styling(fixed_thead = T)%>%
  add_header_above(c("Confirmations" = 1, "Deaths"=1))
  # row_spec(1:5, bold = T, color = "white", background = "#D7261E")

```

### GitHub Log
```{bash gitlog} 
git log --pretty=format:"%nSubject: %s%nAuthor: %aN%nDate: %aD%nBody: %b"
```
